A Simplified Formulation for Molecular Dynamics with Rigid Bodies

### 2016-10-11 rigid_bodies.tex
Início da Correção da Solução Analítica

\newenvironment{smallarray}[1]                          % small arrays
{\null\,\vcenter\bgroup\scriptsize
	\renewcommand{\arraystretch}{0.7}
	\arraycolsep=.13885em
	\hbox\bgroup$\left[\array{@{}#1@{}}}
{\endarray\right]$\egroup\egroup\,\null}

@@

Consider a rigid body composed of $n_p$ individual particles, with each particle $j$ having mass $M_j$ and being located at a Cartesian coordinate $\vt R_j$. The center of mass of this body is located at
\[
\vt r = \frac{1}{m}\sum_{j=1}^{n_p} M_j {\vt R}_j,
\]
where $m = \sum_{j=1}^{n_p} M_j$ is the total mass of the body. The position of each particle $j$ in a space-fixed frame of reference, whose origin coincides with the body's center of mass, is ${\vt \delta}_j = \tr{ [\begin{array}{ccc} {\delta_j}_x & {\delta_j}_y & {\delta_j}_z \end{array}]} = {\vt R}_j - {\vt r}$. In such frame, the symmetric inertia tensor for rotations of the body about the origin is\cite{Goldstein2002}
\[
{\mt I}^\ast = \sum_{j=1}^N M_j \begin{smallarray}{ccc}
{\delta_j}_y^2 + {\delta_j}_z^2 & -{\delta_j}_x {\delta_j}_y & -{\delta_j}_x {\delta_j}_z \\
-{\delta_j}_y {\delta_j}_x & {\delta_j}_x^2 + {\delta_j}_z^2 & -{\delta_j}_y {\delta_j}_z \\
-{\delta_j}_z {\delta_j}_x & -{\delta_j}_z {\delta_j}_y & {\delta_j}_x^2 + {\delta_j}_y^2 \\
\end{smallarray}.
\]

For every non-collinear rigid body (for conciseness, the only case considered here), there exists a particular frame of reference in which its inertia tensor is diagonal. The formulation of the equations of motion in this frame of reference is notably simplified. Conversions from the space-fixed frame to such a body-fixed frame are performed by a rotation matrix $\mt A$ that satisfies the similarity relation\cite{Goldstein2002}
\[
{\mt I} = {\mt A} {\mt I}^\ast \tr{\mt A} = \begin{smallarray}{ccc}
I_1 &   0 &   0 \\
  0 & I_2 &   0 \\
  0 &   0 & I_3 \\
\end{smallarray},
\]
where $I_1$, $I_2$, and $I_3$ (the eigenvalues of ${\mt I}^\ast$) are the principal moments of inertia and the columns of $\tr{\mt A}$ (the corresponding eigenvectors) are the principal axes of rotation. With no loss of generality, we can consider that $I_1 > I_2 > I_3$. Algorithms for obtaining $\mt A$ from $\mt I^\ast$ can be found in the literature (e.g. Ref.~\onlinecite{Kopp2008}). Once this is done, the position of every particle $j$ in the body-fixed frame can be obtained by ${\vt d}_j = {\mt A} {\vt \delta}_j$. Since $\mt A$ represents an orthogonal transformation, a conversion in the opposite direction is done by ${\vt \delta}_j = \tr{\mt A} {\vt d}_j$.\cite{Goldstein2002} This is so because $\mt A \tr{\mt A} = \mt 1_3$, where $\mt 1_k$ represents a $k \times k$ identity matrix. By differentiating both sides of this relation with respect to time, we conclude that ${\mt A} \tr{\dot{\mt A}} = -\dot{\mt A} \tr{\mt A}$, which means that the result of $\mt A \tr{\dot{\mt A}}$ is skew-symmetric. The angular velocity of the body expressed in the body-fixed frame, $\vt \omega$, is related to this product by $\mt S(\vt \omega) = \mt A \tr{\dot{\mt A}}$,\cite{Haug1989} where $\mt S(\cdot)$ is the skew-symmetric operator defined in Appendix \ref{sec:auxiliary_math}  (Eq.~\ref{eq:operator_S}).

@@

Therefore, $\vt q$ is restricted to the surface of a unit hypersphere in $\mathbb{R}^4$, which is itself a three-dimensional domain. Methods are available for computing the unit quaternion related to a given rotation matrix (e.g. Ref.~\onlinecite{Shepperd1978}). Reversely, the matrix $\mt A$ can be directly computed from a given unit quaternion $\vt q$ by\cite{Allen1989,Miller2002}
\[
\label{eq:A_from_q}
\mt A(\vt q) = \begin{smallarray}{ccc}
q_1^2 + q_2^2 - q_3^2 - q_4^2 & 2(q_2 q_3 + q_1 q_4) & 2(q_2 q_4 - q_1 q_3) \\
2(q_2 q_3 - q_1 q_4) & q_1^2 - q_2^2 + q_3^2 - q_4^2 & 2(q_3 q_4 + q_1 q_2) \\
2(q_2 q_4 + q_1 q_3) & 2(q_3 q_4 - q_1 q_2) & q_1^2 - q_2^2 - q_3^2 + q_4^2  
\end{smallarray}.
\]

@@

To conclude this section, we discuss on the possibility of switching between alternative space-fixed frames when describing the motion of a rigid body. This will be useful when we deal with torque-free rotations in Sec.~\ref{sec:free_rotation}. Consider that a body-fixed vector $\vt d$ is, at the same instant $t$, observed as $\vt \delta = \tr{[\mt A(\vt q)]}{\vt d}$ in a space-fixed frame $\mathcal S_q$ and as $\vt \eta = \tr{[\mt A(\vt z)]}{\vt d}$ in another space-fixed frame $\mathcal S_z$. It thus follows that $\vt \eta = \tr{[\mt A(\vt z)]}{\mt A}(\vt q)\vt \delta$, which means that these two reference frames are rotated with respect to each other by a composite matrix $\mt A(\vt g) = \tr{[\mt A(\vt z)]}{\mt A}(\vt q)$. A simple relation between the unit quaternions $\vt g$, $\vt z$, and $\vt q$ is a key result in quaternion algebra.\cite{Dichmann1999} Adapted to our matrix/vector notation, such relation is
\begin{subequations}
\begin{equation}
 \label{eq:comb_rot}
 \vt g = \tr{\left[\begin{array}{cc}\vt z & \mt C(\vt z)\end{array}\right]} \vt q.
\end{equation}

The block matrix above comprises an orthogonal transformation in $\mathbb{R}^4$, which means that the inverse transformation is simply given by
\begin{equation}
\label{eq:inv_comb_rot}
\vt q = \left[\begin{array}{cc}\vt z & \mt C(\vt z)\end{array}\right] \vt g.
\end{equation}
\end{subequations}

As $\mathcal S_q$ and $\mathcal S_z$ are fixed in space, one can conclude that the quaternion $\vt g$ remains constant while both $\vt q$ and $\vt z$ vary in time as the body rotates. In addition, because Eq.~\ref{eq:relation_qdot_omega} was derived for an arbitrary space-fixed frame, it is clear that the equation $\dot{\vt z} = \tfrac{1}{2}{\vt B}(\vt z)\vt \omega$ is equally valid. Therefore, as an alternative to following the evolution of $\vt q$ along a period of time, one can select some convenient unit quaternion $\vt z$ at any instant, compute $\vt g$ using Eq.~\ref{eq:comb_rot}, and then follow the evolution of $\vt z$ from this instant on. Of course, Eq.~\ref{eq:inv_comb_rot} can be employed whenever the value of $\vt q$ is needed afterwards.

@@

The relation between $\vt q$ and $\vt \omega$ (Eq.~\ref{eq:relation_qdot_omega}) is used here to develop a Hamiltonian formalism for the motion of a rigid body. As usual, it can be described by detaching translation and rotation, represented by ${\vt r}(t)$ and ${\vt q}(t)$, respectively. The position of each particle $j$ will vary with time as $\vt r_j(t) = \vt r(t) + \tr{[{\mt A}(\vt q(t))]}\vt d_j$. Several authors have obtained Hamiltonian equations for the rotation of rigid bodies using quaternions as generalized coordinates.\cite{Maciejewski1985, Dichmann1996, Miller2002, Ravishankar2004, Nielsen2012} Here we try to combine convenient features of their formulations. We start by defining the Lagrangian $\mathcal{L} = K - U$ as a function of $\vt r$, $\dot{\vt r}$, $\vt q$, and $\dot{\vt q}$, where $K$ and $U$ are the kinetic and potential energies of the body, respectively. Since $K = \frac{1}{2} m \tr{\dot{\vt r}} \dot{\vt r} + \frac{1}{2} \tr{\vt \omega} \mt I \vt \omega$,\cite{Goldstein2002} we can resort to Eq.~\ref{eq:relation_qdot_omega} to write
\[
\mathcal{L} = \frac{1}{2} m \tr{\dot{\vt r}} \dot{\vt r} + 2 \tr{\dot{\vt q}} \mt B \mt I \tr{\mt B} \dot{\vt q} - U(\vt r, \vt q).
\]

@@

It is not possible Eq.~\ref{eq:EDO_system}

Method based on work of Kosenko\cite{Kosenko_1998} and Celledoni \textit{et al}.\cite{Celledoni_2008} and also related to the work of van Zon \textit{et al}.\cite{vanZon2007}.

Considering that $I_1 > I_2 > I_3$, the prefactors of the angular momenta $\omega_1$, $\omega_2$, and $\omega_3$ are, respectively,
\begin{align}
\mt A = \left[\begin{array}{ccc}
\frac{\omega_1^0}{\sqrt{1 - (\omega_2^0/\lambda_1)^2}} & 0 & 0 \\
0 & \min(\lambda_1,\lambda_3) & 0 \\
0 & 0 & \frac{\omega_3^0}{\sqrt{1 - (\omega_2^0/\lambda_3)^2}}
\end{array}\right], \\
\end{align}
\begin{align}
A_1 = \frac{\omega_1^0}{\sqrt{1 - (\omega_2^0/\lambda_1)^2}}, \\
A_2 = \min(\lambda_1,\lambda_3), \; \text{and} \\
A_3 = \frac{\omega_3^0}{\sqrt{1 - (\omega_2^0/\lambda_3)^2}},
\end{align}
where
\begin{equation}
\lambda_1 = \sqrt{\frac{L^2 - 2 I_3 K_r}{I_2(I_2 - I_3)}} \;\; \text{and} \;\; \lambda_3 = \sqrt{\frac{2 I_1 K_r - L^2}{I_2(I_1 - I_2)}}.
\end{equation}

The solution is
\begin{subequations}
\begin{align}
\omega_1(t) &= \begin{cases}
A_1\text{cn}(a t + b,k) \quad \text{if} \quad \lambda_1 < \lambda_3 \\
A_1\text{dn}(a t + b,k) \quad \text{if} \quad \lambda_1 > \lambda_3
\end{cases} \\
\omega_2(t) &= A_2 \text{sn}(a t + b,k) \\
\omega_3(t) &= \begin{cases}
A_3\text{dn}(a t + b,k) \quad \text{if} \quad \lambda_1 < \lambda_3 \\
A_3\text{cn}(a t + b,k) \quad \text{if} \quad \lambda_1 > \lambda_3
\end{cases}
\end{align}
\end{subequations}
where
\begin{align*}
&a = -\frac{(I_1-I_3)A_1A_3}{I_2 A_2},\\
&b = \text{sn}^{-1}\left(\frac{\omega_{2,0}}{A_2},k\right),\quad\text{and}\\
&k = \frac{A_2}{\text{max}(\lambda_1,\lambda_3)}
\end{align*}

During a torque-free rotational motion, the angular momentum vector $\vt L$ remains constant in any space-fixed frame of reference.

In the body-fixed frame, although the

>>>>

\newcommand{\sn}{\text{sn}}                             % Jacobi function sn
\newcommand{\cn}{\text{cn}}                             % Jacobi function cn
\newcommand{\dn}{\text{dn}}                             % Jacobi function dn

\newenvironment{smallarray}[1]                          % small arrays
{\null\,\vcenter\bgroup\scriptsize
	\renewcommand{\arraystretch}{1.5}
	\arraycolsep=.13885em
	\hbox\bgroup$\left[\array{@{}#1@{}}}
{\endarray\right]$\egroup\egroup\,\null}

@@

Consider a rigid body composed of $n_p$ individual particles, with each particle $j$ having mass $M_j$ and being located at a Cartesian coordinate $\vt R_j$. The center of mass of this body is located at
\begin{equation*}
\vt r = \frac{1}{m}\sum_{j=1}^{n_p} M_j {\vt R}_j,
\end{equation*}
where $m = \sum_{j=1}^{n_p} M_j$ is the total mass of the body. The position of each particle $j$ in a space-fixed frame of reference, whose origin coincides with the body's center of mass, is ${\vt \delta}_j = \tr{ [\begin{array}{ccc} {\delta_j}_x & {\delta_j}_y & {\delta_j}_z \end{array}]} = {\vt R}_j - {\vt r}$. In such frame, the symmetric inertia tensor for rotations of the body about the origin is\cite{Goldstein2002}
\begin{equation*}
{\mt I}^\ast = \sum_{j=1}^N M_j \begin{smallarray}{ccc}
{\delta_j}_y^2 + {\delta_j}_z^2 & -{\delta_j}_x {\delta_j}_y & -{\delta_j}_x {\delta_j}_z \\
-{\delta_j}_y {\delta_j}_x & {\delta_j}_x^2 + {\delta_j}_z^2 & -{\delta_j}_y {\delta_j}_z \\
-{\delta_j}_z {\delta_j}_x & -{\delta_j}_z {\delta_j}_y & {\delta_j}_x^2 + {\delta_j}_y^2 \\
\end{smallarray}.
\end{equation*}

For every non-collinear rigid body (for conciseness, the only case considered here), there exists a particular frame of reference in which its inertia tensor is diagonal. The formulation of the equations of motion in this frame of reference is notably simplified. Conversions from the space-fixed frame to such a body-fixed frame are performed by a rotation matrix $\mt A$ that satisfies the similarity relation\cite{Goldstein2002}
\begin{equation*}
{\mt I} = {\mt A} {\mt I}^\ast \tr{\mt A} = \begin{smallarray}{ccc}
I_1 &   0 &   0 \\
0 & I_2 &   0 \\
0 &   0 & I_3 \\
\end{smallarray},
\end{equation*}
where $I_1$, $I_2$, and $I_3$ (the eigenvalues of ${\mt I}^\ast$) are the principal moments of inertia and the columns of $\tr{\mt A}$ (the corresponding eigenvectors) are the principal axes of rotation. With no loss of generality, we can consider that $I_1 \geq I_2 \geq I_3$. Algorithms for obtaining $\mt A$ from $\mt I^\ast$ can be found in the literature (e.g. Ref.~\onlinecite{Kopp2008}). Once this is done, the position of every particle $j$ in the body-fixed frame can be obtained by ${\vt d}_j = {\mt A} {\vt \delta}_j$. Since $\mt A$ represents an orthogonal transformation, a conversion in the opposite direction is done by ${\vt \delta}_j = \tr{\mt A} {\vt d}_j$.\cite{Goldstein2002} This is so because $\mt A \tr{\mt A} = \mt 1_3$, where $\mt 1_k$ represents a $k \times k$ identity matrix. By differentiating both sides of this relation with respect to time, we conclude that ${\mt A} \tr{\dot{\mt A}} = -\dot{\mt A} \tr{\mt A}$, which means that the result of $\mt A \tr{\dot{\mt A}}$ is skew-symmetric. The angular velocity of the body expressed in the body-fixed frame, $\vt \omega$, is related to this product by $\mt S(\vt \omega) = \mt A \tr{\dot{\mt A}}$,\cite{Haug1989} where $\mt S(\cdot)$ is the skew-symmetric operator defined in Appendix \ref{sec:auxiliary_math}  (Eq.~\ref{eq:operator_S}).

@@

Therefore, $\vt q$ is restricted to the surface of a unit hypersphere in $\mathbb{R}^4$, which is itself a three-dimensional domain. Methods are available for computing the unit quaternion related to a given rotation matrix (e.g. Ref.~\onlinecite{Shepperd1978}). Reversely, the matrix $\mt A$ can be directly computed from a given unit quaternion $\vt q$ by\cite{Allen1989,Miller2002}
\begin{equation*}
\label{eq:A_from_q}
\mt A(\vt q) = \begin{smallarray}{ccc}
q_1^2 + q_2^2 - q_3^2 - q_4^2 & 2(q_2 q_3 + q_1 q_4) & 2(q_2 q_4 - q_1 q_3) \\
2(q_2 q_3 - q_1 q_4) & q_1^2 - q_2^2 + q_3^2 - q_4^2 & 2(q_3 q_4 + q_1 q_2) \\
2(q_2 q_4 + q_1 q_3) & 2(q_3 q_4 - q_1 q_2) & q_1^2 - q_2^2 - q_3^2 + q_4^2  
\end{smallarray}.
\end{equation*}

@@

To conclude this section, we discuss on the possibility of switching between alternative space-fixed frames when describing the motion of a rigid body. This will be useful when we deal with torque-free rotations (Sec.~\ref{sec:free_rotation}). Consider that a body-fixed vector $\vt d$ is simultaneously observed as $\vt \delta = \tr{\mt A\!}(\vt q){\vt d}$ in a space-fixed frame $\mathcal S_q$ and as $\vt \xi = \tr{\mt A\!}(\vt z){\vt d}$ in another space-fixed frame $\mathcal S_z$. It follows that $\vt \xi = \mt A(\vt g)\vt \delta$, where the composite rotation matrix $\mt A(\vt g) = \tr{\mt A\!}(\vt z){\mt A}(\vt q)$ denotes the relative orientation between $\mathcal S_q$ and $\mathcal S_z$. One key result in quaternion algebra\cite{Dichmann1999} is a simple relation that exists between $\vt q$, $\vt z$, and $\vt g$, which can be adapted to our notation as $\vt g = \tr{[\begin{array}{cc}\vt z & \mt C(\vt z)\end{array}]} \vt q$. Note that both $\vt q$ and $\vt z$ vary while the body rotates, but the quaternion $\vt g$ remains constant because $\mathcal S_q$ and $\mathcal S_z$ are fixed in space. Moreover, as Eq.~\ref{eq:relation_qdot_omega} was derived for an arbitrary space-fixed frame, it is also true that $\dot{\vt z} = \tfrac{1}{2}{\vt B}(\vt z)\vt \omega$. In conclusion, as an alternative to tracking the evolution of $\vt q(t)$ after an instant $t_0$ in which $\vt q = \vt q_0$, one can select some other convenient unit quaternion $\vt z_0$ and then track the evolution of $\vt z(t)$, starting from $\vt z(t_0) = \vt z_0$. Then, whenever $\vt q$ is needed, one can simply make 
\begin{equation}
\label{eq:relation_q_z}
\vt q = [\begin{array}{cc}\vt z & \mt C(\vt z)\end{array}] \tr{[\begin{array}{cc}\vt z_0 & \mt C(\vt z_0)\end{array}]} \vt q_0,
\end{equation}
which is correct because these block matrices represent orthogonal transformations in $\mathbb{R}^4$.

@@

The relation between $\vt q$ and $\vt \omega$ (Eq.~\ref{eq:relation_qdot_omega}) is used here to develop a Hamiltonian formalism for the motion of a rigid body. As usual, it can be described by detaching translation and rotation, represented by ${\vt r}(t)$ and ${\vt q}(t)$, respectively. The position of each particle $j$ will vary with time as $\vt r_j(t) = \vt r(t) + \tr{[{\mt A}(\vt q(t))]}\vt d_j$. Several authors have obtained Hamiltonian equations for the rotation of rigid bodies using quaternions as generalized coordinates.\cite{Maciejewski1985, Dichmann1996, Miller2002, Ravishankar2004, Nielsen2012} Here we try to combine convenient features of their formulations. We start by defining the Lagrangian $\mathcal{L} = K - U$ as a function of $\vt r$, $\dot{\vt r}$, $\vt q$, and $\dot{\vt q}$, where $K$ and $U$ are the kinetic and potential energies of the body, respectively. Since $K = \frac{1}{2} m \tr{\dot{\vt r}} \dot{\vt r} + \frac{1}{2} \tr{\vt \omega} \mt I \vt \omega$,\cite{Goldstein2002} we can resort to Eq.~\ref{eq:relation_qdot_omega} to write
\begin{equation*}
\mathcal{L} = \frac{1}{2} m \tr{\dot{\vt r}} \dot{\vt r} + 2 \tr{\dot{\vt q}} \mt B \mt I \tr{\mt B} \dot{\vt q} - U(\vt r, \vt q).
\end{equation*}

@@

It is not possible to analytically integrate the equations of motion for a general potential $U(\vt r,\vt q)$. Nonetheless, closed-form solutions for Eqs.~\ref{eq:EDO_q} and \ref{eq:EDO_pi} are known for the particular case of torque-free rotation (i.e. when $\vt \tau = \vt 0$). Free rotations are very useful in Molecular Dynamics as part of numerical integration methods based on splitting strategies, which is the topic addressed in Sec.~\ref{sec:symplectic}. As shown in Appendix~\ref{sec:derivation_euler}, Eq.~\ref{eq:EDO_pi} results in the well-known Euler equation $\mt I \dot{\vt \omega} + \vt \omega \times (\mt I \vt \omega) = \mt A \vt \tau$, which is completely decoupled from Eq.~\ref{eq:EDO_pi} when $\vt \tau = \vt 0$. It is a known fact that both the rotational energy $K_r = \frac{1}{2}\tr{\vt \omega}{\mt I}{\vt \omega}$ and the space-fixed angular momentum vector $\vt L$ (whose squared norm is $L^2 = \tr{\vt \omega}{\mt I}^2{\vt \omega}$) are integrals of motion in a free rotation. This allows us to analytically solve the Euler equation for the angular velocity vector $\vt \omega$. Here we present the solution for an asymmetric top ($I_1 > I_2 > I_3$) in a form that aims to facilitate computer implementation, inspired by the derivations presented in Refs.~\onlinecite{Landau1976} and \onlinecite{Thorn2012}. From an initial condition $\vt \omega(0) = \vt \omega_0 = \tr{[\omega_1^0 \; \omega_2^0 \; \omega_3^0]}$, such solution starts by computing $K_r = \frac{1}{2}\tr{\vt \omega_0}{\mt I}{\vt \omega}_0$ and $L = (\tr{\vt \omega_0}{\mt I}^2{\vt \omega_0})^{1/2}$, followed by evaluating a vector of prefactors defined as
\begin{equation*}
\vt a = \begin{smallarray}{c}
\frac{\omega_1^0}{\sqrt{1 - \left(\omega_2^0/\lambda_1\right)^2}} \\
\min(\lambda_1,\lambda_3) \\
\frac{\omega_3^0}{\sqrt{1 - (\omega_2^0/\lambda_3)^2}}
\end{smallarray},
\; \text{where} \;
\begin{cases}
\lambda_1 = \sqrt{\frac{L^2 - 2 I_3 K_r}{I_2(I_2 - I_3)}} \\
\lambda_3 = \sqrt{\frac{2 I_1 K_r - L^2}{I_2(I_1 - I_2)}}
\end{cases}.
\end{equation*}

Note that the value of $a_2$ depends on a comparison between $\lambda_1$ and $\lambda_3$. This also occurs for $\vt \omega$, whose functional form is
\begin{subequations}
\begin{align}
\vt \omega(t) &= \begin{smallarray}{c}
a_1\text{cn}(\omega_p t + \epsilon, k^2) \\
a_2\text{sn}(\omega_p t + \epsilon, k^2) \\
a_3\text{dn}(\omega_p t + \epsilon, k^2)
\end{smallarray} \quad \text{if} \quad \lambda_1 < \lambda_3 \quad \text{or} \\
\vt \omega(t) &= \begin{smallarray}{c}
a_1\text{dn}(\omega_p t + \epsilon, k^2) \\
a_2\text{sn}(\omega_p t + \epsilon, k^2) \\
a_3\text{cn}(\omega_p t + \epsilon, k^2)
\end{smallarray} \quad \text{if} \quad \lambda_1 > \lambda_3,
\end{align}
\end{subequations}
where $\text{sn}(u,m)$, $\text{cn}(u,m)$, and $\text{dn}(u,m)$ are the Jacobi elliptic functions,\cite{Byrd_1971} $k = {a_2}/{\text{max}(\lambda_1,\lambda_3)}$ is the elliptic modulus, and the constants $\omega_p$ and $\epsilon$ are given by
\begin{equation*}
\omega_p = -\frac{(I_1 - I_3)a_1 a_3}{I_2 a_2} \quad \text{and} \quad
\epsilon = \text{sn}^{-1}\left(\frac{\omega_2^0}{a_2}, k^2\right).
\end{equation*}

In order to compute the inverse function $\text{sn}^{-1}$, one can use its relation with an elliptic integral in Carlson form, \cite{Carlson_1977} $\text{sn}^{-1}(x,m) = x R_F(1-x^2,1-mx^2,1)$. In this work, we employ the routines available in Ref.~\onlinecite{Galassi_2009} for numerically evaluating Jacobi functions and Carlson integrals.

Once $\vt \omega$ is known as a function of time, we can solve Eq.~\ref{eq:EDO_q} with initial condition $\vt q(0) = \vt q_0$. The method we propose here, which is closely related to the one pioneered by Kosenko\cite{Kosenko_1998} and extended by Celledoni \textit{et al}.,\cite{Celledoni_2008} relies on the factorization in Eq.~\ref{eq:factorization_of_A} to facilitate the task. During a free rotation, any space-fixed representation of the angular momentum vector will remain constant. Thus, instead of solving Eq.~\ref{eq:EDO_q} directly for obtaining $\vt q$, we solve an analogous equation for obtaining another unit quaternion $\vt z$ specially chosen so that $\tr{[\mt A(\vt z)]}\mt I\vt \omega = \tr{[\begin{array}{ccc} L & 0 & 0 \end{array}]}$. In other words, we look for an alternative space-fixed frame whose first axis is aligned with the angular momentum vector.

\begin{equation*}
\mt B(\vt z)\mt I\vt \omega = \mt C(\vt z)\begin{smallarray}{c}
L \\ 0 \\ 0
\end{smallarray}
\end{equation*}

By making ${\mt A}(\vt z) = \mt A(\vt z)\mt C(\vt z)$

factorizing $\mt A(\vt z)$ and applying Eq.~\ref{eq:product_B_vector}, we obtain a homogeneous linear system
\begin{equation*}
\Big(L \hat{\vt C}_1 - \sum_{j=1}^3 I_j \omega_j \hat{\vt B}_j\Big) \vt z = \vt 0,
\end{equation*}
where $\hat{\vt C}_1$ and $\hat{\vt B}_k$ are the permutation matrices introduced in Appendix~\ref{sec:auxiliary_math}. Hence, besides having unit norm, the quaternion $\vt z$ will always lie in the null space of a matrix $\mt G = L \hat{\vt C}_1 - \sum_{j=1}^3 I_j \omega_j \hat{\vt B}_j$, which is guaranteed to exist because $\det \mt G = (L^2 - \tr{\vt \omega}{\mt I}^2{\vt \omega})^2 = 0$. Such a null space turns out to be a moving hyperplane in $\mathbb{R}^4$ defined by the orthogonal basis vectors
\begin{equation*}
\vt v_1(t) = \begin{smallarray}{c}
\frac{I_3\omega_3(t)}{L-I_1\omega_1(t)} \\
\frac{I_2\omega_2(t)}{L-I_1\omega_1(t)} \\
1 \\
0
\end{smallarray}
\quad \text{and} \quad
\vt v_2(t) = \begin{smallarray}{c}
\frac{-I_2\omega_2(t)}{L-I_1\omega_1(t)} \\
\frac{I_3\omega_3(t)}{L-I_1\omega_1(t)} \\
0 \\
1
\end{smallarray}.
\end{equation*}

At all times, therefore, $\vt z$ will be a linear combination of $\vt v_1$ and $\vt v_2$. As these basis vectors are known functions of time, the problem of determining $\vt z(t)$ reduces to finding out how the two coefficients of the combination vary with time. Examination of $\vt v_1$ and $\vt v_2$ above makes it clear that these coefficients coincide with the entries $z_3$ and $z_4$ of $\vt z$, in such a way that
\begin{equation}
\label{eq:linear_comb_z}
\vt z(t) = z_3(t) \vt v_1(t) + z_4(t) \vt v_2(t).
\end{equation}

Note that we are free to choose any unit quaternion in the hyperplane defined by $\vt v_1(0)$ and $\vt v_2(0)$ as the initial condition $\vt z(0) = \vt z_0$. In other words, we can select any space-fixed frame whose first axis aligns with the angular momentum vector. For convenience, we choose
\begin{equation*}
\vt z_0 = \frac{\vt v_1(0)}{\|\vt v_1(0)\|} = \sqrt{\frac{L - I_1\omega_1^0}{2L}} \vt v_1(0).
\end{equation*}

By substituting Eq.~\ref{eq:linear_comb_z} into $\dot{\vt z} = \frac{1}{2}{\mt B}(\vt z){\vt \omega}$, which is the equivalent of Eq.~\ref{eq:EDO_q} for $\vt z$, we obtain a vector of four differential equations containing $z_3$ and $z_4$, from which we pick the third and fourth ones, which are linearly independent. After replacing $(I_2 - I_3)\omega_2\omega_3 = I_1\dot{\omega}_1$ and $I_1\omega_1^2 + I_2\omega_2^2 + I_3\omega_3^2 = 2 K_r$, followed by defining
\begin{equation*}
\mt M(t) = \begin{smallarray}{rr} f(t) & -g(t) \\ g(t) & f(t) \end{smallarray} \; \text{with} \;
\begin{cases}
f(t) = \frac{-I_1 \dot{\omega}_1(t)}{L - I_1\omega_1(t)} \\
g(t) = \frac{2K_r - L\omega_1(t)}{L - I_1\omega_1(t)}
\end{cases},
\end{equation*}
we can represent these equations in matrix form as
\begin{equation*}
\begin{smallarray}{c} \dot{z}_3 \\ \dot{z}_4 \end{smallarray} = \frac{1}{2} {\mt M}(t) \begin{smallarray}{c} z_3 \\ z_4 \end{smallarray}.
\end{equation*}

It is fortunate that the product $\mt M(t_a)\mt M(t_b)$ commutes for any $t_a$ and $t_b$, since this is a sufficient condition for a solution to be obtained in terms of matrix exponential $\exp[\frac{1}{2}\int_0^t \mt M(\tau)d\tau]$. Such solution is
\begin{equation*}
\begin{smallarray}{c} z_3 \\ z_4 \end{smallarray} = \sqrt{\tfrac{L-I_1\omega_1(t)}{L-I_1\omega_1^0}} \begin{smallarray}{rr} \cos \phi(t) & -\sin \phi(t) \\ \sin \phi(t) & \cos \phi(t) \end{smallarray} \begin{smallarray}{c} z_3^0 \\ z_4^0 \end{smallarray},
\end{equation*}
where $\phi(t) = \frac{1}{2} \int_0^t g(\tau)d\tau$. Note that this consists in a counterclockwise rotation, followed by a scaling.

\begin{align*}
z_3(t) &= \sqrt{\tfrac{L-I_1\omega_1(t)}{2L}} \cos \phi(t) \\
z_4(t) &= \sqrt{\tfrac{L-I_1\omega_1(t)}{2L}} \sin \phi(t)
\end{align*}

\begin{equation*}
\phi(t) = \frac{L}{2 I_1\omega_p}\left\{(A_\phi + 1)\omega_p t + A_\phi[h(t) - h(0)]\right\},
\end{equation*}
where $A_\phi = \frac{2 I_1 K_r - L^2}{L^2 - I_1^2 a_1^2}$ and $B_\phi = \frac{\max(\lambda_1,\lambda_3)}{\sqrt{\lambda_1^2 + \eta \lambda_3^2}}$, $C_\phi = \frac{\lambda_1\lambda_3}{a_3 \sqrt{\lambda_1^2 + \eta \lambda_3^2}}$. The function $h(t)$ is defined as
\begin{align*}
h(t) &= \Theta\left(\frac{\omega_2(t)}{a_2},-\frac{a_2^2 }{\lambda_1^2}\eta,k^2\right) + \\
&+ \frac{I_1}{L} \frac{a_1 a_3}{a_2} C_\phi \arctan\left(C_\phi\frac{\omega_3(t)}{\omega_2(t)}\right)
\end{align*}

But $\frac{a_1 a_3}{a_2} = - \frac{I_2}{I_1 - I_3} \omega_p$. Thus,
\begin{align*}
h(t) &= \Theta\left(\frac{\omega_2(t)}{a_2},-\frac{a_2^2 }{\lambda_1^2}\eta,k^2\right) \\
&- \frac{I_1 \omega_p}{L}\frac{I_2 C_\phi}{I_1 - I_3} \arctan\left(C_\phi\frac{\omega_3(t)}{\omega_2(t)}\right)
\end{align*}

Rewriting $g(t)$:
\begin{equation*}
g(t) = \frac{L}{I_1} + \frac{2 I_1 K_r - L^2}{L I_1 \omega_p}\frac{\omega_p}{1 + \alpha[\omega_1(t)/a_1]},
\end{equation*}
where $I_1 a_1/L$. By defining $u = \omega_p \tau + \epsilon$, we have that $du = \omega_p d\tau$. Therefore, if $\lambda_1 < \lambda_3$:
\begin{equation*}
\phi(t) = \frac{L}{2I_1}t + \frac{2 I_1 K_r - L^2}{2L I_1 \omega_p}\int\limits_\epsilon^{\omega_p t + \epsilon} \frac{1}{1 + \alpha \cn u}du
\end{equation*}

Otherwise, if $\lambda_3 < \lambda_1$:
\begin{equation*}
\phi(t) = \frac{L}{2I_1}t + \frac{2 I_1 K_r - L^2}{2L I_1 \omega_p}\int\limits_\epsilon^{\omega_p t + \epsilon} \frac{1}{1 + \alpha \dn u}du
\end{equation*}

From Ref.~\onlinecite{Byrd_1971}, we deduce that
\begin{align*}
\int \frac{1}{1 + \alpha \cn u}du = (\eta + 1) \Bigg[\Pi(\sn u,-\eta,k^2) + \\
+ \frac{\alpha}{\sqrt{k^2+\eta}}\arctan\left(\frac{\dn u}{\sn u \sqrt{k^2+\eta}}\right)\Bigg]
\end{align*}

Given that $\cn(u,k^2) = \dn(ku,1/k^2)$, $\dn(u,k^2) = \cn(ku,1/k^2)$, and $\sn(ku,1/k^2) = k\sn(u,k^2)$
\begin{align*}
\int \frac{1}{1 + \alpha \dn u}du = \frac{\eta + 1}{k} \Bigg[\Pi(k\sn u,-\eta,1/k^2) + \\
+ \frac{\alpha}{\sqrt{1/k^2+\eta}}\arctan\left(\frac{\cn u}{k \sn u \sqrt{1/k^2+\eta}}\right)\Bigg]
\end{align*}
where $\eta = \frac{\alpha^2}{1-\alpha^2}$. Moreover, as $\Pi(x,y,z) = k\Pi(x/k,yk^2,zk^4)$:
\begin{align*}
\int \frac{1}{1 + \alpha \dn u}du = (\eta + 1) \Bigg[\Pi(\sn u,-k^2\eta,k^2) + \\
+ \frac{\alpha}{k\sqrt{1/k^2+\eta}}\arctan\left(\frac{\cn u}{k \sn u \sqrt{1/k^2+\eta}}\right)\Bigg]
\end{align*}

\begin{widetext}
\begin{equation*}
\phi(t) = \frac{L}{2 I_1\omega_p}\left\{(A_\phi + 1)\omega_p t + A_\phi [h(t) - h(0)]\right\},
\end{equation*}
\begin{equation}
h(t) = \Theta\left(\frac{\omega_2(t)}{a_2},-\frac{a_2^2 }{\lambda_1^2}\eta,k^2\right) + \alpha B_\phi \arctan\left(B_\phi\frac{\omega_3(t)/a_3}{\omega_2(t)/a_2}\right)
\end{equation}
where $A_\phi = \frac{2 I_1 K_r - L^2}{L^2}(\eta + 1)$ and $B_\phi = \frac{\max(\lambda_1,\lambda_3)}{\sqrt{\lambda_1^2 + \eta \lambda_3^2}}$
